Follow-up Samples

### 2017-01-13 submission/manuscript_outline_20161024.Rmd
add updated pdf

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between stool bacterial microbiome and CRC.   The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.   
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`.  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.  Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.   
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

**Treatment and Time Differences**

There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).

>>>>

**Background:** Colorectal cancer (CRC) continues to be a worldwide health problem with early detection being used as a key component in mitigating deaths due to the disease. Previous research suggests a link between stool bacterial microbiome and CRC.   The overall objective was to investigate the changes in the bacterial microbiome after surgery in patients with lesion (i.e. adenoma or carcinoma).  Specifically, we wanted to identify what within the community was different within those undergoing surgical removal of lesion.  We also investigated the use of the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build models to either classify individuals as having a lesion or whether, based on the bacterial microbiome, the sample could be classified correctly as before or after surgery.   
**Results:** Adenoma individual's bacterial microbiome were more similar to their pre-surgery sample then those with carcinoma (P-value = `r format(change_summary['thetayc', 'Pvalue'], digits = 3)`) and this was also reflected in FIT as well (P-value = `r format(change_summary['fit', 'Pvalue'], scientific = TRUE, digits = 3)`).  There was no significant difference in any indivdiual OTU between samples before and after surgery (P-value > `r format(min(filter(all_OTU_pair_wilcox, analysis == "ALL")[, 'BH_corrected']), digits = 3)`).  A model with a total of `r length(lesion_top10_varsMDA$variable)` variables was able to classify lesion with an AUC range of `r format(max(reduced_lesion_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_lesion_AUC_run_summary$ROC), digits = 3)` while the model to classify samples as before and after had `r length(IF_top10_varsMDA$variable)` with an AUC range of `r format(max(reduced_IF_AUC_run_summary$ROC), digits = 3)` to `r format(min(reduced_IF_AUC_run_summary$ROC), digits = 3)` for `r length(reduced_lesion_AUC_run_summary$ROC)` 20 repeated 10-fold cross-validated runs.  Both models had a significant decrease in the positive probability of a lesion between individual's before versus after surgery samples (P-value = `r format(filter(all_probs_paired_wilcox, model_type == "red_lesion" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)` and `r format(filter(all_probs_paired_wilcox, model_type == "red_IF" & comparison =="lesion")[, "BH_correction"], scientific = TRUE, digits = 3)`).  In total there were `r length(common_vars_summary$otu)` OTUs that were common to both models and were mostly commensals with largest representation from OTUs belonging to `r common_vars_summary["Otu000005", "lowest_ID"]`, `r common_vars_summary["Otu000012", "lowest_ID"]`, `r common_vars_summary["Otu000082", "lowest_ID"]`, and `r paste(gsub("_unclassified", "", common_vars_summary["Otu000006", "lowest_ID"]))`.   
**Conclusions:** Our data suggests that treatment not only significantly reduces the probability of having a colonic lesion but also causes detectable changes in the bacterial microbiome.  Further surveillance of these individuals will enable us to determine whether models such as the one we present here can also be used to predict recurrence of colorectal cancer.

@@

**Treatment and Time Differences**
There was no difference in the amount of change in positive probability for either the full or reduced lesion model for either chemotherapy (P-value = `r format(chemo_rad_summary['chemo_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['chemo_red_les', 'bh'], digits = 3)`) or radiation therapy (P-value = `r format(chemo_rad_summary['rad_les', 'bh'], digits = 3)` and `r format(chemo_rad_summary['rad_red_les', 'bh'], digits = 3)`). Although the initial follow up model was similar there was a significant decrease in positive probability for those treated with chemotherapy (P-value = `r format(chemo_rad_summary['chemo_IF', 'bh'], scientific = TRUE, digits = 3)` and `r format(chemo_rad_summary['chemo_red_IF', 'bh'], scientific = TRUE, digits = 3)`Time of follow up sample from initial sampling, did not have a significant difference between adenoma and carcinoma (uncorrected P-value = `r format(time_pvalues["thetayc_change", "uncorrected_p_value"], digits = 3)`).
